Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight Data

### 2011-04-20 extended_catalog/paper/extended_catalog.tex
Adding more plots

\documentclass[12pt]{article}
\usepackage{fullpage}
\usepackage{graphicx}

\title{Search for Spatially Extended Sources Using Two Years of Flight
Data}

@@

\maketitle

\abstract{
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that were done to validate this tool.
We then apply this tool to test all sources in the 2FGL Two Year Catalog
for extension.\cite{2FGL} We then report on the discovery of XXX new
extended sources.
}

@@

@@

Performing morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent point spread function (PSF) and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

{\em pointlike} optimizes the analysis of extended sources in two ways.
First, for extended sources which are radially symmetric, the convolution
of the source shape with our parameterization of the PSF can be performed
semi-analytically. Generally, at a given energy the observed 
photon distribution is
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A' 
\end{equation}
where $I_\text{src}$ is the extended soruce shape and the PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma}
\end{equation}
Here, $\sigma$ and $\gamma$ are parameterize the $PSF$.
If the extended source is radially symmetric, then
$I_\text{src} (\vec r') = I_\text{src} (r')$ and the angular part of the
integral can be done analytically, leaving a simpler one dimensional
integral:
\begin{equation}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma
  ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right)
\end{equation}
Not only is this convolution formula faster, but it has the advantage
that it more accurately reduces to the actual PSF shape as the source's
size decreases.  Decreasing the numerical difference between a small extended
source and the PSF is important when searching for smaller extended sources
and when determining the statistical significance of a detection.

@@

In going from fitting a point source to fitting a simple extended source such as
a disk, the source's width is the only degree of freedom added to the model. According to
Wilk's Theorem, the distribution of $\text{TS}_\text{ext}$ for no signal (when there is a true
point source) should follow a $\chi^2$ distribution with one degree of freedom\cite{Wilks_Theorem}.
Wilk's Theorem only holds when the likelihood is maximized simultaneoulsy
fitting all of the parameters of model. It is for this reason why {\em pointlike}
goes through such great effort to simultaneously fit the position, extension,
and spectral paramters of a source.

@@

It seems plausable that a similar situation could happen for when 
testing a source for extension. Half of the time, statistical flucutations
might make the observed counts narrower thant the point spread function
and the likelhood would not improve by fitting the source with an extended
hypothesis. We tested this hypothesis using a dedicated monte carlo study.

\begin{figure}
\includegraphics{mc_plots/ts_ext_emin_1000.pdf}
% this plot came from /u/gl/lande/work/fermi/extended_catalog/monte_carlo/ts_ext/v3/plot.py
% using data from /nfs/slac/g/ki/ki03/lande/extended_catalog/monte_carlo/ts_ext/v3_emin_1000/merged.hdf5
\caption{Lalla}
\end{figure}

@@

@@

@@

>>>>

\documentclass{emulateapj}

\usepackage{fullpage}
\usepackage{graphicx}

\shorttitle{Search for Extended LAT Sources}

@@

\title{Search for Spatially Extended Fermi-LAT Sources Using Two Years of Flight
Data}

\author{
J.~Lande\altaffilmark{3}, 
\altaffiltext{3}{W. W. Hansen Experimental Physics Laboratory, Kavli Institute for Particle Astrophysics and Cosmology, Department of Physics and SLAC National Accelerator Laboratory, Stanford University, Stanford, CA 94305, USA}
}

\begin{abstract}
We present a new method for analyzing spatially extended sources with
the Fermi Large Area Telescope (LAT), the primary science instrument
on the {\em Fermi Gamma-ray Space Telescope (Fermi)}.  We provide a
series of Monte Carlo studies that were done to validate this tool.
We then apply this tool to test all sources in the 2FGL Two Year Catalog
for extension.\cite{2FGL} We then report on the discovery of XXX new
extended sources.
\end{abstract}

@@

The Large Area Telescope (LAT) is a pair conversion telescope on the
Fermi Gamma-Ray Space Telescope that has been surveying the gamma ray
sky since June 2008.  Fermi is particuarly well suited for surveying
the entire Gamma-ray sky since it has a broad energy coverage (20MeV
and 300GeV), a wide field of view (~2.4 sr) and large effective area
($\approx 8000 \text{cm}^2$ at > 1GeV).

The LAT has been essential in understanding a variety of astrophysical
sources including Active Galactic Nuclei, Pulsars, and Supernova Remnants.
Many of the astrophysical sources seen by the Fermi LAT, espeically
galactic Supernova Remnants and Pulsar Wind Nebulae, can be resolved at
other wavelengths and could potentially be resolved by the LAT.

Resolving spatially extended GeV sources is important for studying a
variety of science topics. The most important use of spatial morphology
is in source identification. Many of the sources found in GeV sources

Nevertheless, morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are

studying supernova remnants, 
pulsar wind nebulae, 
 * galaxies,  
 * potentially dark matter or new physics.

@@

Performing morphological studies of sources in the GeV energy range using
the Fermi LAT is challenging primarily because of the significantly energy
dependent PSF and because of systematic errors in the
galactic diffuse emission. The Fermi PSF varies from over $XXXXX^\circ$
at 100MeV to $XXXXX^\circ$ at 100GeV and so the higher energy photons are
significantly more important for studying the shape of sources. One could
ideally look just at the highest energy photons, but the trade off is that
for typical sources many more source events are expected at low energy.
For this reason, it is especially important to use as much of the data
as possible.

@@

{\em pointlike} optimizes the analysis of extended sources in two ways.
First, for extended sources which are radially symmetric, the convolution
of the source shape with our parameterization of the PSF can be performed
semi-analytically. Generally, at a given energy the observed 
photon distribution is
\begin{equation}
  \text{PDF}(\vec r) = \int  \text{PSF}(|\vec r - \vec r'|)I_\text{src}(\vec r') d A' 
\end{equation}
where $I_\text{src}$ is the extended soruce shape and the PSF is parameterized by
\begin{equation}
  \text{PSF}(\vec r) = 
  \frac{1}{2\pi\sigma^2}
  \left(1-\frac{1}{\gamma}\right)
  \left(1+\frac{u}{\gamma}\right)^{-\gamma}
\end{equation}
Here, $\sigma$ and $\gamma$ are parameterize the $PSF$.
If the extended source is radially symmetric, then
$I_\text{src} (\vec r') = I_\text{src} (r')$ and the angular part of the
integral can be done analytically, leaving a simpler one dimensional
integral:
\begin{multline}
  \text{PDF}(u)= \int_0^\infty dv
  I_\text{src}(v) 
  \left(\frac{\gamma-1}{\gamma}\right)
  \left( \frac{\gamma}{\gamma + u + v}\right)^\gamma \\
  \times ~_2F_1 \left(\gamma/2,\frac{1+\gamma}{2},1,\frac{4uv}{(\gamma+u+v)^2}\right)
\end{multline}
Not only is this convolution formula faster, but it has the advantage
that it more accurately reduces to the actual PSF shape as the source's
size decreases.  Decreasing the numerical difference between a small extended
source and the PSF is important when searching for smaller extended sources
and when determining the statistical significance of a detection.

@@

In going from fitting a point source to fitting a simple extended source such as
a disk, the source's width is the only degree of freedom added to the model. According to
Wilk's Theorem, the distribution of $\text{TS}_\text{ext}$ for no signal (when there is a true
point source) should follow a $\chi^2_1$ distribution with one degree of freedom\cite{Wilks_Theorem}.
Wilk's Theorem only holds when the likelihood is maximized simultaneoulsy
fitting all of the parameters of model. It is for this reason why {\em pointlike}
goes through such great effort to simultaneously fit the position, extension,
and spectral paramters of a source.

@@

It seems plausable that a similar situation could happen for when testing
a source for extension. Half of the time, statistical flucutations might
make the observed counts narrower thant the point spread function and
the likelhood would not improve by fitting the source with an extended
hypothesis. We tested this hypothesis below using a dedicated monte
carlo study.

@@

We performed a dedicated monte carlo study to validate pointlike for
studying spatially extended sources and also to demonstrate that we can
use $\text{TS}_\text{ext}$ as a test of the statistical significance
of a source's extension.  The monte carlo study involved simulating point
sources of varying spectral models on top of a simulated background. The
simulated point source was then fit using pointlike as both an extended
source and a point source and $\text{TS}_\text{ext}$ was calculated.

Fermi LAT data was simulated using the publically avaliabable Science
Tools package {\em gtobssim}\cite{GTOBSSIM_CITATION}. The sources were
simulated with a powerlaw spectral model with 6 Were given 6 $>100\text{MeV}$
fluxes ranging from $3\times 10^{-9} ph/cm^2/s$ to $10^{-6} ph/cm^2/s$.
The sources were then simulated with spectral index ($\gamma$) values of
1.5, 2, 2.5, and 3.  These values were picked to probe a representative
range measurable values.  For particular spectral indices, only the
fluxes which would create a statistically significant source detection
($\text{TS}>25$) where selected, since we are only interested in
testing for extension sources with a significant extent. The sources were
simulated on top of an isotropic powerlaw background with a Sreekumar-like
spectrum ($>\text{100MeV}$ flux of 1.5e-5 and spectral index 2.1).\cite{Sreekumar
et al. ApJ 494 pag 523 1998}

The monte carlo simulation was perforormed over a one year time interval using
a default rocking profile and an assumed livetime fraction of 0.8.
The reconstruction was perfromed with events from an energy range of
1GeV to 100GeV using 4 energy bins per decade and using the Pass 7 Version 6
source IRFs to be consistent with the following analysis.

For most of the spectral models, over 10,000 statistically independent simulations were performed.
For fainter sources, some of the simulated sources, statistical flucatuations
caused the source to be not significant and they were excluded from the following analysis.
Table~\ref{ts_ext_num_sims} shows the parametesr used in the simualtion,
the number of simulations perfromed, and the average test statistic value.

\begin{table}
  \begin{centering}
\begin{tabular}{ | r | r | r | r | }
\hline
$\gamma$ & flux ($ph/cm^2/s$) & $N_\text{sims}$ & $<\text{TS}>$ \\
\hline
  1.5 &          $10^{-6}$ &           11405 &  90861 \\
      &  $3\times 10^{-7}$ &           11490 &  21922 \\
      &          $10^{-7}$ &           11101 &   5765 \\
      &  $3\times 10^{-8}$ &           10269 &   1262 \\
      &          $10^{-8}$ &            9941 &    306 \\
      &  $3\times 10^{-9}$ &           10427 &     61 \\
\hline
    2 &          $10^{-6}$ &           11632 &  22025 \\
      &  $3\times 10^{-7}$ &           11622 &   4886 \\
      &           $1^{-7}$ &           11115 &   1159 \\
      &  $3\times 10^{-8}$ &           10503 &    222 \\
      &           $1^{-8}$ &            9893 &     50 \\
\hline
  2.5 &          $10^{-6}$ &           11522 &   4722 \\
      &  $3\times 10^{-7}$ &           11445 &    948 \\
      &           $1^{-7}$ &           10966 &    202 \\
      &  $3\times 10^{-8}$ &            8227 &     41 \\
\hline                                                
    3 &          $10^{-6}$ &           11264 &    949 \\
      &  $3\times 10^{-7}$ &           10986 &    172 \\
      &           $1^{-7}$ &            7855 &     40 \\
\hline
\end{tabular}
\caption{List of the spectral models used to the monte carlo simulation of
{\em pointlike}.$N_\text{sims}$ is the number of monte carlo simluations
which produce a significant source detection.The final column is the
average test statistic that was found during the monte carlo fit of
the source. Only simluations which were deemed significant ($\text{TS}>25$)
were kept.}
\label{ts_ext_num_sims}
  \end{centering}
\end{table}

The results of the monte carlo simulation are shown in
figure~\ref{ts_ext}.  The cumulative density function (CDF) is plotted
against $\text{TS}_\text{ext}$ For reference, the CDF for $\chi^2_1/2$,
suggested by Wilk's Theorem, is also plotted for comparison. Overall,
there is good agreement between Wilk's theorem, although the agreement
is not perfect.  Possible reasons for departure from Wilk's theorem may
include pointlike ignoring energy dispersion, which would generally change
the shape as a function of energy, numerical innacuracies in evaluating
the convolution algorithm, or possibly problems with the fitting algorithm
getting stuck in false minimum. It is important to emphasize that if
our emperical curve lies to the left of the curve we assume to estimate
significance, we would underestimate the statistical significance of
our result. Since this is the case for most of our simluations, we are
reasonable confident that $\sqrt{\text{TS}_\text{ext}}$ can be safely
used as a measure of the statistical significance of a source's extension.

@@

We use another monte carlo study to determine the Fermi LAT's sensitivity to
spatially extended sources.

@@

\begin{figure}
  \begin{center}
    \includegraphics{mc_plots/ts_ext_emin_1000.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/monte_carlo/ts_ext/v3/plot.py
    % using data from /nfs/slac/g/ki/ki03/lande/extended_catalog/monte_carlo/ts_ext/v3_emin_1000/merged.hdf5
    \end{center}
    \caption{Lalla}\label{ts_ext}
  \end{figure}

\begin{figure}
  \begin{center}
    \includegraphics{mc_plots/sensitivity.pdf}
    % this plot came from /u/gl/lande/work/fermi/extended_catalog/monte_carlo/sensitivity/v8/merged_plot.py
    % using data from /nfs/slac/g/ki/ki03/lande/extended_catalog/monte_carlo/sensitivity/v7/merged.hdf5 (>1GeV)
    % using data from /nfs/slac/g/ki/ki03/lande/extended_catalog/monte_carlo/sensitivity/v8/merged.hdf5 (>100MeV)
    \end{center}
    \caption{Lalla}\label{sensitivity_ext}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J1628.6-2419c}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_1000_v1/1FGL_J1628.6-2419c/v2/source_1FGL_J1628.6-2419c.pdf
    \caption{Lalla}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J1554.0-5345c}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_1000_v1/1FGL_J1554.0-5345c/v2/source_1FGL_J1554.0-5345c.pdf
    \caption{Lalla}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J0823.3-4248}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_1000_v1/1FGL_J0823.3-4248/v2/source_1FGL_J0823.3-4248.pdf
    \caption{Lalla}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J2020.0+4049}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_10000_v1/1FGL_J2020.0+4049/v2/source_1FGL_J2020.0+4049.pdf
    \caption{Lalla}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_MRFH0254}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_10000_v1/MRFH0254/v2/source_MRFH0254.pdf
    \caption{Lalla}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J1837.5-0659c}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_10000_v1/1FGL_J1837.5-0659c/v2/source_1FGL_J1837.5-0659c.pdf
    \caption{1FGL J1837.5-0659c}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J1636.4-4737c}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_10000_v1/1FGL_J1636.4-4737c/v2/source_1FGL_J1636.4-4737c.pdf
    \caption{1FGL J1636.4-4737c}
  \end{figure}

  \begin{figure}
    \begin{center}
      \includegraphics[type=pdf,ext=.pdf,read=.pdf]{source_plots/source_1FGL_J1613.6-5100c}
    \end{center}
    % this plot came from 
    % /nfs/slac/g/ki/ki03/lande/extended_catalog/2FGL/v10/spectral_emin_10000_v1/1FGL_J1613.6-5100c/v2/source_1FGL_J1613.6-5100c.pdf
    \caption{Lalla}
  \end{figure}
